---
title: "Exercise 1 - Getting started with data science"
author: "Ng Jian Long"
output: html_document
---

Assignments for the course focus on practical aspects of the concepts covered in the lectures. Assignments are based on the material covered in James et al. You will start working on the assignment in the lab sessions after the lectures, but may need to finish them after class.  If you do not get through the whole assignment, do not worry!

## 1. Working with a sample dataset

This exercise relates to the `College` data set, which can be found in the file `College.csv` on the website for the main course textbook (James et al 2013) http://faculty.marshall.usc.edu/gareth-james/ISL/data.html. It contains a number of variables for 777 different universities and colleges in the US.   

The variables are:    
* `Private` : Public/private indicator  
* `Apps` : Number of applications received  
* `Accept` : Number of applicants accepted  
* `Enroll` : Number of new students enrolled  
* `Top10perc` : New students from top 10% of high school class   
* `Top25perc` : New students from top 25% of high school class  
* `F.Undergrad` : Number of full-time undergraduates  
* `P.Undergrad` : Number of part-time undergraduates  
* `Outstate` : Out-of-state tuition  
* `Room.Board` : Room and board costs  
* `Books` : Estimated book costs  
* `Personal` : Estimated personal spending  
* `PhD` : Percent of faculty with Ph.D.'s  
* `Terminal` : Percent of faculty with terminal degree  
* `S.F.Ratio` : Student/faculty ratio  
* `perc.alumni` : Percent of alumni who donate  
* `Expend` : Instructional expenditure per student  
* `Grad.Rate` : Graduation rate  

Before reading the data into R, it can be viewed in Excel or a text editor, if you find that convenient.

### 1.1 Load the data

Use the `read.csv()` function to read the data into `R`. Call the loaded data `college`. Make sure that you have the directory set to the correct location for the data.  You can load this in R directly from the website, using:

```{r, quietly = TRUE}
library(ISLR)

#Loading Data
college <- read.csv("College.csv")
```

### 1.2 View the data

Look at the data using the `View()` function. You should notice that the first column is just the name of each university. We don't really want R to treat this as data. However, it may be handy to have these names for later. Try the following commands:

```{r}
row.names <- college[, 1]
rownames(college) <- row.names
View(college)
```

You should see that there is now a `row.names` column with the name of each university recorded. This means that R has given each row a name corresponding to the appropriate university. R will not try to perform calculations on the row names. However, we still need to eliminate the first column in the data where the names are stored. Try

```{r}
#Remove the first column
college <- college[, -1] 
View(college)

#Convert Binary variable
college$Private_num <- ifelse(college$Private == "Yes", 1, 0)
View(college)

mean(college$Private_num)
```

Now you should see that the first data column is `Private`. Note that another column labeled `row.names` now appears before the `Private` column. However, this is not a data column but rather the name that `R` is giving to each row.

### 1.3 Try some operations

a.  Use the `summary()` function to produce a numerical summary of the variables in the data set.

```{r}
library(tidyverse, quietly = TRUE)
library(dplyr, quietly = TRUE)

#Summary statistics
summary(college)

#Summary statistics in list form
lapply(college, summary)

#Column names
names(college)

#Summary statistics in matrix form
sapply(college[,-c(1,3:4)],summary)
```
  
b.  Use the `pairs()` function to produce a scatterplot matrix of the first ten columns or variables of the data. Recall that you can reference the first ten columns of a matrix `A` using `A[,1:10]`.

```{r}
#Scatterplot matrix
pairs(college[,2:10])
```
   
c.  Use the `plot()` function to produce side-by-side boxplots of `Outstate` versus `Private`.

```{r}
#Boxplots
Outstate <- college$Outstate
Private <- college$Private
Name <- c("Outstate","Private")
boxplot(Outstate ~ Private, data = college)
```

d.  Create a new qualitative variable, called `Elite`, by *binning* the `Top10perc` variable. We are going to divide universities into two groups based on whether or not the proportion of students coming from the top 10% of their high school classes exceeds 50%.

```{r}
#Creating the Elite variable
Elite <- rep("No", nrow(college))

#Mark "Yes" if >50% of the student coming from top 10% of their high school classes
Elite[college$Top10perc > 50] <- "Yes"

#Convert the Elite variable into numeric values
class(Elite) #The variable is of character type
Elite <- as.factor(Elite) #Convert it into numeric

#Combine both together
college <- data.frame(college, Elite)
```

Use the `summary()` function to see how many elite universities there are. Now use the `plot()` function to produce side-by-side boxplots of `Outstate` versus `Elite`.

```{r} 
summary(college$Elite)

Outstate <- college$Outstate
Elite1 <- college$Elite
Label <- c("Outstate","Elite")
boxplot(Elite, Outstate, names = Label,horizontal = TRUE)

plot(Outstate,Elite,xlab="Elite Universities",ylab="Tuition in $")
```

Continue exploring the data, and provide a brief summary of what you discover.

```{r} 
View(College)

#Further exploration
acceptance_rate <- college$Accept / college$Apps

{
plot(college$Outstate,college$Grad.Rate)
plot(acceptance_rate, college$S.F.Ratio)
plot(college$Top10perc,college$Grad.Rate)
}

{
cat(paste("College with the most students in the top 10% class: ", row.names(college)[which.max(college$Top10perc)], '\n'))
cat(paste("College with the lowest acceptance rate: ", row.names(college)[which.min(acceptance_rate)], '\n'))
cat(paste("College with the highest acceptance rate: ", row.names(college)[which.max(acceptance_rate)], '\n'))
cat(paste("College with the highest graduation rate: ", row.names(college)[which.max(college$Grad.Rate)], '\n'))
}
```
### 2. Working with the automotive dataset

This exercise involves the `Auto` data set available as `Auto.csv` from the website for the main course textbook James et al. http://faculty.marshall.usc.edu/gareth-james/ISL/data.html. Make sure that the missing values have been removed from the data. You should load that dataset as the first step of the exercise. 

```{r} 
#Load data
Auto <- read.csv("Auto.csv")

#Remove NA values
Auto <- na.omit(Auto)

#Summary statistics
summary(Auto)
```

(a) Which of the predictors are quantitative, and which are qualitative?
**Quantitative - mpg, cylinder, displacement, horsepower, weight, acceleration, year.**
**Qualitative - origin, Name.**
    
(b) What is the *range* of each quantitative predictor? You can answer this using the `range()` function.

```{r} 
#Excluding quantitative variables
Auto1 <- Auto[,-c(8:9)]

apply(Auto1, 2, range)
```

(c) What is the mean and standard deviation of each quantitative predictor?

```{r} 
mean<- apply(Auto1, 2, mean)
sd <- apply(Auto1, 2, sd)

cbind(mean = mean, sd = sd)
```

(d) Now remove the 10th through 85th observations. What is the range, mean, and standard deviation of each predictor in the subset of the data that remains?

```{r} 
Auto2 <- Auto1[-c(10:85),]

range <- apply(Auto2, 2, range)
mean <- apply(Auto2, 2, mean)
sd <- apply(Auto2, 2, sd)

tbl <- rbind(range, mean, sd)
rownames(tbl) <- c("Min", "Max", "Mean", "sd")
tbl
```

(e) Using the full data set, investigate the predictors graphically, using scatterplots or other tools of your choice. Create some plots highlighting the relationships among the predictors. Comment on your findings.

```{r} 
#Use Auto1 as Auto1 excludes qualitative variables
library(purrr, quietly = TRUE)
library(corrplot, quietly = TRUE)

corr <- cor(Auto1) #Calculate correlation between variables
high_value <- apply(abs(corr) >= 0.8, 2, any) #Extract correlation greater than 0.8
corrplot(corr[high_value, high_value], method = "number") #plot variables with corr > 0.8

```

**The variable "mpg", "displacement", "cylinders", "horsepower" and "weight" are highly correlated with each other.**
    
(f) Suppose that we wish to predict gas mileage (`mpg`) on the basis of the other variables. Do your plots suggest that any of the other variables might be useful in predicting `mpg`? Justify your answer.

```{r}
#Prediction
Predict <- lm(mpg ~ cylinders+displacement+horsepower+weight+acceleration+year+origin, data=Auto)
summary(Predict)
```

**The t-values suggests "displacement", "weight", "year" and "origin" are significant in predicting "mpg". The remaining variables are not significant in predicting the model.**
    
```{r}
fit <- lm(mpg ~ displacement+weight+year+origin, data=Auto)
summary(fit)
```
    
### Further materials

- [Data Camp R tutorials](https://www.datacamp.com/courses/free-introduction-to-r)
- [Data Camp R Markdown tutorials](https://www.datacamp.com/courses/reporting-with-r-markdown). You can complete the free first chapter. 
